---
title: "Case Study"
author: "Group 4"
date: "2025-11-09"
output: powerpoint_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(gridExtra)
library(plm)
library(fect)
```

## R Markdown

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Load and Wrangle Data
```{r data_import}
# Create link to dataset
url <- 'https://raw.githubusercontent.com/pmwest7/case_study/refs/heads/main/synthetic_sales.csv'

# Read file into dataframe
data <- read.csv(url)

```

```{r glimpse}
data |> glimpse()
```

```{r summary}
data |> summary()
```

```{r}
# Add new variable converting group assignments to binary 0/1
data <- data |>
  mutate(assignment = as.numeric(group == 'Treatment'))

data |> head()
```
## Exploratory Data Analysis (EDA)
```{r barplot}
group_size <- ggplot(data, mapping = aes(x = group, label= ..count..)) +
  geom_bar(fill='#1170aa') +
  geom_text(stat = 'count', vjust=-0.5, size = 5) +
  ylim(0, 350)

group_size

```

```{r hist_boxplot}

# histogram of April sales per hour
apr_sales_per_hour <- ggplot(data, mapping = aes(x = apr_sph)) +
  geom_histogram(binwidth=100, color = '#ffffff', fill='#1170aa') +
  xlim(0, 2500) +
  ylim(0, 80)

# histogram of May sales per hour
may_sales_per_hour <- ggplot(data, mapping = aes(x = may_sph)) +
  geom_histogram(binwidth=100, color = '#ffffff', fill='#fc7d0b') +
   xlim(0, 2500) +
  ylim(0, 80)

apr_summary <- ggplot(data, mapping = aes(x = factor(group), y = apr_sph)) +
  geom_boxplot() +
  ylim(0, 2500)

may_summary <- ggplot(data, mapping = aes(x = factor(group), y = may_sph)) +
  geom_boxplot() +
  ylim(0, 2500)

grid.arrange(apr_sales_per_hour, may_sales_per_hour, apr_summary, may_summary, ncol=2, nrow=2)
```
## Parallel Trends
```{r panel_data }
# Create a new dataframe for use in evaluating the pre-treatment period for both the treated and untreated groups
pre_kU <- data |>
  select(assignment, apr_sph) |> # append group values and April sales/hour to the new dataframe
  rename(sales_per_hr = apr_sph) |> # append column to standardize sales per hour variable label
  mutate(time=0, # append a new variable, time, set equal to 1 to indicate the pre-period
         did_assign = assignment, # append a new variable and copy assignment values
         assignment = 0)# update assignment to 0 to indicate pre-treatment assignment

# Create a new dataframe for use in evaluating the post-treatment period for both the treated and untreated groups         
post_kU <- data |>
  select(assignment, may_sph) |> # append group values and May sales/hour to the new dataframe
  rename(sales_per_hr = may_sph) |> # append column to standardize sales per hour variable label
  mutate(time=1, # append a new variable, time, set equal to 2 to indicate the post-period
         did_assign = assignment)# append a new variable indicating the true group assignment

# Combine the pre- and post-treatment observations
kU_2x2_DD <- bind_rows(pre_kU, post_kU) |> 
  mutate(id = rep(data$sp_id, 2)) |> # replicate the employee id number to match pre- and post- treatment records for employees
  mutate(did_assign = factor(did_assign),
         id = factor(id)) # encode variables as factors (categories)

```

```{r}
ggplot(kU_2x2_DD, aes(x = time, y = sales_per_hr, color=did_assign)) +
      geom_smooth(alpha=0.1, method="lm", formula=y ~ x) + 
      labs(title='Evaluation of Sales Per Hour (SPH) From April to May',
           x = 'Period',
           y = 'Sales/Hour (SPH)',
           color = 'Group') + 
  theme_bw()

```

## Difference-in-Differences
```{r}

# Use Panel Data Estimators to evaluate differences in mean within units between the two periods

panel_df <- pdata.frame(kU_2x2_DD, index=c('id', 'time'))

standard_did <- plm(sales_per_hr ~ assignment,
                    data = panel_df,
                    model = 'within',
                    effect = 'twoways')

standard_did |> summary()

```

## Fixed Effects Counterfactual Estimators
Not statistically significant
confidence interval contains zero
p-value > 0.05
```{r fixed effect}
# https://yiqingxu.org/packages/fect/

counter_est <- fect(sales_per_hr ~ assignment,
                    data = kU_2x2_DD,
                    Y = 'sales_per_hr', # outcome
                    D = 'assignment', # treatment
                    index = c('id', 'time'),
                    seed = 7890,
                    se = TRUE)

counter_est

```

```{r fect_plot}
plot(counter_est, count = FALSE, return.test = TRUE) # returns plots and test statistics

```

```{r}


```