---
title: "Case Study"
author: "Group 4"
date: "2025-11-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, incliude = FALSE)

if (!requireNamespace('pwr', quietly = TRUE)) {
    install.packages('pwr')
}

library(tidyverse)
library(gridExtra)
library(pwr)
library(plm)
library(fect)
```

## Define Outcome
$H_0:$ There is no difference in the average sales per hour between the treatment group, `k`, and control group, `U`.

$H_a:$ There is a difference in the average sales per hour between the treatment group, `k`, and control group, `U`.

## Load and Wrangle Data
```{r data_import}
# Create link to dataset
url <- 'https://raw.githubusercontent.com/pmwest7/case_study/refs/heads/main/synthetic_sales.csv'

# Read file into dataframe
data <- read.csv(url)

```
## Examine data structure
```{r glimpse}
data |> glimpse()
```
## Examine Summary Statistics
```{r summary}
data |> summary()
```
## Transform Data
```{r}
# Add new variable converting group assignments to binary 0/1
data <- data |>
  mutate(assignment = as.numeric(group == 'Treatment'))

# Create segments for control group
data <- data |>
mutate(segment =case_when(group == 'Control' ~ ntile(apr_sph, 4), group == 'Treatment' ~ segment))

data |> head()
```


## Exploratory Data Analysis (EDA)
```{r barplot}
group_size <- ggplot(data, mapping = aes(x = group, label= ..count..)) +
  geom_bar(fill = c('#fc7d0b', '#1170aa')) +
  geom_text(stat = 'count', vjust=-0.5, size = 5) +
  ylim(0, 350)

group_size

```

```{r}
group_segments <- data |>
  group_by(group = as.factor(group), segment) |>
  summarize(count = n(), .groups = 'drop')

group_segments

 ggplot(group_segments, mapping = aes(x = segment, y = count, fill= group)) +
  geom_col() +
   scale_fill_manual(values = c('#fc7d0b', '#1170aa')) +
   labs(title = 'Segmentation Distribution by Group') +
   geom_text(aes(label = count, y=count), vjust=-0.5, size = 3) + 
   facet_wrap(~ group)
```


```{r hist_boxplot}

# histogram of April sales per hour
apr_sales_per_hour <- ggplot(data, mapping = aes(x = apr_sph)) +
  geom_histogram(binwidth=100, color = '#ffffff', fill='#1170aa') +
  xlim(0, 2500) +
  ylim(0, 80)

# histogram of May sales per hour
may_sales_per_hour <- ggplot(data, mapping = aes(x = may_sph)) +
  geom_histogram(binwidth=100, color = '#ffffff', fill='#fc7d0b') +
   xlim(0, 2500) +
  ylim(0, 80)

apr_summary <- ggplot(data, mapping = aes(x = factor(group), y = apr_sph)) +
  geom_boxplot() +
  ylim(0, 2500)

may_summary <- ggplot(data, mapping = aes(x = factor(group), y = may_sph)) +
  geom_boxplot() +
  ylim(0, 2500)

grid.arrange(apr_sales_per_hour, may_sales_per_hour, apr_summary, may_summary, ncol=2, nrow=2)
```

## Power Analysis $1 - \beta$
### Cohen's d
The small control group size impacts the model's ability to find statistical significance, resulting in a high probability of failing to reject a false $H_0$. In this case 82% chance.
```{r}
# Explore impact of imbalance between treatment and control groups

# Standard deviation of control
control_diff <- data$may_sph[data$assignment == 0]- data$apr_sph[data$assignment == 0]

sd_0 <- sd(control_diff, na.rm = TRUE)

# Standard deviation of treatment
treat_diff <- data$may_sph[data$assignment == 1]- data$apr_sph[data$assignment == 1]

sd_1 <- sd(treat_diff, na.rm = TRUE)

pooled_sd <- sqrt((sd_0^2 + sd_1^2)/2)

# Control mean
mu_0 <- mean(control_diff, na.rm = TRUE)

# Treatment mean
mu_1 <- mean(treat_diff, na.rm = TRUE)

# Estimation of effect size
cohensd <- (mu_1 - mu_0)/pooled_sd

# Sample size of treatment and control group
n_0 <- sum(data$assignment == 0)
n_1 <- sum(data$assignment == 1)

power_of_test <- pwr.t2n.test(n1 = 24, n2 = 326, d = cohensd, sig.level = 0.05, alternative = "two.sided")

power_of_test
```
## Parallel Trends
```{r panel_data }
# Create a new dataframe for use in evaluating the pre-treatment period for both the treated and untreated groups
pre_kU <- data |>
  select(assignment, apr_sph, apr_quantity, apr_price, apr_rts, segment) |> # append group values and April sales/hour to the new dataframe
  rename(sales_per_hr = apr_sph, quantity = apr_quantity, price = apr_price, rts = apr_rts) |> # append column to standardize sales per hour variable label
  mutate(time=0, # append a new variable, time, set equal to 1 to indicate the pre-period
         did_assign = assignment, # append a new variable and copy assignment values
         assignment = 0)# update assignment to 0 to indicate pre-treatment assignment


# Create a new dataframe for use in evaluating the post-treatment period for both the treated and untreated groups         
post_kU <- data |>
  select(assignment, may_sph, may_quantity, may_price, may_rts, segment) |> # append group values and May sales/hour to the new dataframe
  rename(sales_per_hr = may_sph, quantity = may_quantity, price = may_price, rts = may_rts) |> # append column to standardize sales per hour variable label
  mutate(time=1, # append a new variable, time, set equal to 2 to indicate the post-period
         did_assign = assignment)# append a new variable indicating the true group assignment


# Combine the pre- and post-treatment observations
kU_2x2_DD <- bind_rows(pre_kU, post_kU) |> 
  mutate(id = rep(data$sp_id, 2)) |> # replicate the employee id number to match pre- and post- treatment records for employees
  mutate(did_assign = factor(did_assign),
         id = factor(id),
         segment = factor(segment)) # encode variables as factors (categories)

```

```{r}
ggplot(kU_2x2_DD, aes(x = time, y = sales_per_hr, color=did_assign)) +
      geom_smooth(alpha=0.1, method="lm", formula=y ~ x) + 
      labs(title='Evaluation of Sales Per Hour (SPH) From April to May',
           x = 'Period',
           y = 'Sales/Hour (SPH)',
           color = 'Group') + 
      scale_color_manual(values = c('#fc7d0b','#1170aa')) +
  theme_bw()

```

## Difference-in-Differences (aggregate)
```{r}

# Use Panel Data Estimators to evaluate differences in mean within units between the two periods

panel_df <- pdata.frame(kU_2x2_DD, index=c('id', 'time'))

standard_did <- plm(sales_per_hr ~ assignment,
                    data = panel_df,
                    model = 'within',
                    effect = 'twoways')

standard_did |> summary()

```
## Fixed Effects Counterfactual Estimators
Not statistically significant
- confidence interval contains zero
- p-value > 0.05
```{r fixed effect}

fect_est <- fect(sales_per_hr ~ assignment,
                    data = kU_2x2_DD,
                    Y = 'sales_per_hr', # outcome
                    D = 'assignment', # treatment
                    index = c('id', 'time'),
                    seed = 7890,
                    se = TRUE)

fect_est

```

```{r fect_plot}
plot(fect_est,
     count = FALSE,
     return.test = TRUE) # returns plots and test statistics

```

```{r}
plot(fect_est,
     plot.ci = "0.95",
     show.points = FALSE)
```

```{r}
percent_diff_0 <- mu_0/mean(data$apr_sph[data$assignment == 0])
percent_diff_1 <- mu_1/mean(data$apr_sph[data$assignment == 1])

sprintf('The original findings are consistent with the standard DID esitmator. The change in the control group was %.2f, and the change in the treatment group was %.2f.', X <- percent_diff_0*100, y <- percent_diff_1*100)

```

## Fixed Effects Counterfactual Estimators (segmented)
Not statistically significant
- confidence interval contains zero
- p-value > 0.05
```{r fixed effect segment}
fect_est_segment <- fect(sales_per_hr ~ assignment,
                    data = kU_2x2_DD,
                    Y = 'sales_per_hr', # outcome
                    D = 'assignment', # treatment
                    group = 'segment',
                    index = c('id', 'time'),
                    seed = 7890,
                    se = TRUE)

fect_est_segment$est.group.att

```

## Fixed Effects Counterfactual Estimators (segmented - quantity)
Statistically significant results for Segments 1 and 2, the traditionally low performers
-  confidence interval does not contain zero
- p-value < 0.05
```{r fixed effect quantity}

fect_quantity <- fect(quantity ~ assignment,
                    data = kU_2x2_DD,
                    Y = 'quantity', # outcome
                    D = 'assignment', # treatment
                    group = 'segment',
                    na.rm = TRUE,
                    index = c('id', 'time'),
                    seed = 7890,
                    se = TRUE
                    )

fect_quantity$est.group.att

```

## Fixed Effects Counterfactual Estimators (segmented - price)

```{r fixed effect price}

fect_price <- fect(price ~ assignment,
                    data = kU_2x2_DD,
                    Y = 'price', # outcome
                    D = 'assignment', # treatment
                    group = 'segment',
                    index = c('id', 'time'),
                    seed = 7890,
                    se = TRUE)

fect_price$est.group.att/2 # Why is this ATT double the value in the case study? Divided by two to make the values match. The other variables calculated to the the correct amount

```

## Fixed Effects Counterfactual Estimators (segmented - rts)
statistically significant
- confidence interval does not contain zero
- p-value < 0.05
```{r fixed effect rts}

fect_rts <- fect(rts ~ assignment,
                    data = kU_2x2_DD,
                    Y = 'rts', # outcome
                    D = 'assignment', # treatment
                    group = 'segment',
                    index = c('id', 'time'),
                    seed = 7890,
                    se = TRUE)

fect_rts$est.group.att

```

```{r}
treat_segments <- data |>
  select(segment, apr_sph, may_sph, assignment) |>
  filter(assignment == 1)|>
  group_by(segment) |>
  summarize(mean_difference = mean(may_sph - apr_sph), mu1 = mean(may_sph)) |>
  mutate(ATT = (mean_difference/mu1)*100) |>
  column_to_rownames(var='segment')

 
control_segments <- data |>
  select(segment, apr_sph, may_sph, assignment) |>
  filter(assignment == 0)|>
  group_by(segment) |>
  summarize(mean_difference = mean(may_sph - apr_sph), mu0 = mean(may_sph)) |>
  mutate(ATT = (mean_difference/mu0)*100) |>
  column_to_rownames(var='segment')

treat_segments
control_segments

treat_segments - control_segments

```

## References:
Bounthavong, M. (2021, December 29). *Sample size estimation and power analysis*. RPubs.com. https://rpubs.com/mbounthavong/sample_size_power_analysis_R

Cohen, J. (1988). Statistical power analysis for the behavioral sciences. Taylor & Francis Group. https://ebookcentral.proquest.com/lib/unc/detail.action?docID=1192162

Ismay, c., Kim, A.Y. (2025, September 16). *A moderndive into R and the tidyverse*. https://moderndive.com/index.html

Xu, Y. (2022). *fect*. [User Manual].GitHub.  https://yiqingxu.org/packages/fect/